### 2014-12-12 paper.tex
Formatting nits and trailing whitespace

  The very simplest metrics for compactness simply measure the length and width of the district as compared to that of a circle or rectangle of the same area.  There are several variations on this method, but they are all fairly crude.  Other metrics compare the area of the district to the area of the smallest circumscribing circle, or the largest inscribed circle, or to the convex hull of the district -- the smallest convex shape containing the entire district.  Niemi~et~al.~\cite{niemi} list a number of variations on these metrics.  An alternate method of computing dispersion is by computing the moment of inertia of the area of the district.  The concept is borrowed from physics, and is defined as
  \[I = \iint ((x-x_0)^2 + (y-y_0)^2) dx dy \]
  where $(x_0, y_0)$ is the center of mass of the district (weighted by area).  More dispersed districts have higher moments of inertia, because the average square distance of a point in the district to the center is larger.  In order to ensure that the metric does not depend on the size of the state, and to make it range from 0 to 1 with 0 being the least compact, we in fact use the measure $A/\sqrt{2\pi I}$.  This metric seems to have been first proposed by Kaiser.~\cite{kaiser}

@@

  Lastly, since ultimately it is not the geography that matters, but the people that live in it, we can modify some of these metrics to take into account population.  Those that compare the area of the district to that of the smallest circumscribing circle or that of the convex hull of the district may simply compare population instead of area, as proposed by Hofeller and Grofman.~\cite{hofeller}.  Alternately, instead of computing the area-weighted moment of inertia, one can compute the population-weighted moment of inertia, which is defined as
  \[I_P = \iint ((x-x_0)^2 + (y-y_0)^2) \rho dx dy \]
  where $\rho$ is the population density and $(x_0, y_0$ is the center of mass of the district, now weighted by population.  In practice, since we do not know the location of each individual in the district, we may simply assume that each person lives at the center of the census tabulation block in which they were counted.  Then the formula becomes
  \[I_P = \sum_{b} ((x_b-x_0)^2 + (y_b-y_0)^2) P_b\]
  where the sum is taken over census tabulation blocks $b$, with $(x_b, y_b)$ the center of mass (weighted by area) of the census block, and $P_b$ its population.  In order to make the metric less dependent on population, we can use the metric $AP_d/(2\pi I_P)$; this does not necessarily normalize the metric to be between 0 and 1, but at least makes it independent of the total area and total population.  Using moment of population to measure district compactness was first proposed by Weaver.~\cite{weaver}

@@

    \subsection{Partisan Symmetry and Vote Efficiency}
 While geographic compactness is a widely-considered metric for compactness, others have focused on actual results for determining unacceptable redistricting plans.  Over the last few years, the courts have increasingly rejected a variety of rationales cited to claim illegal gerrymandering, including ``partisan intent'',  ``minority party entrenchment'', and ``lack of proportionality''.  However, in \textit{League of United Latin American Citizens (LULAC) v. Perry, Governor of Texas et al.}, (2006), the majority recognized that it would accept a lack of \textit{symmetry} as an acceptable argument for determining unconstitutional gerrymandering. \\

  A given plan has symmetry if ``similary-situated parties are treated equally'', thus expecting that an even split in votes across two parties would have an even split in seats. Furthermore, if a plaintiff challenging a redistricting plan can demonstrate partisan intent and asymmetry, it would be sufficient for the burden of proof to shift to the state.  Nevertheless, while acknowledging the viability of symmetry as a metric to determine a discriminatory impact on voters, the court failed to establish an accepted standard or test to determine illegal gerrymandering.  Existing measures at the time, as calculated by social scientists, used hypothetical models to predict how election outcomes would turn out, given certain shifts in the electorate and certain vote shares in a given election. The court rejected the notion of using hypothetical models for a standard of constitutionality. \cite{LULAC} \\

  With this in mind,  as an alternative to other existing metrics, two legal scholars recently proposed measuring \textit{vote efficiency} as a way to quantify partisan symmetry and partisan gerrymandering \cite{stephanopoulos}.  Vote efficiency is a measure of how well a given vote applied towards improving political representation.  In the case that a voter's desired candidate wins, any vote beyond what is necessary to win (in the simple case, 50\% + 1 vote) is wasted.  In the case when a voter's desired candidate loses, all votes for the losing candidate are wasted. Redistricting plans seeking a partisan bias will lead to fewer wasted votes for the favored party, while increasing wasted votes for the disadvantaged party. \\

@@

\subsection{Redistricting Processes}
Another key aspect when discussing gerrymandering is the political processes involved in redistricting.  The Constitution provides states with the ability to create their own rules in how districts are drawn, with the provisions that states will ensure balanced district populations, and that districts are redrawn at least every 10 years \cite{LULAC}.\\

Currently, there are five general types of processes, some of which overlap.  The first, and most common, is when redistricting is done by the \textit{legislature}.    The legislature proposes plans and seeks to pass a bill for approval.  Sometimes, this may require a supermajority, or may be veto-proof, but typically requires simple majority approval from both the upper and lower houses and from the governor.  The second is the \textit{advisory commission} model, where a group of people, usually appointed officials, gather together to make non-binding recommendations to the legislature.  The third is the \textit{backup commission}, where failure of the legislature to pass a new plan triggers the activation of a predetermined committee to create an alternative.  The fourth is a \textit{political commission}, where elected officials often serve as members of a redistricting commission.  This is the sole legislative process for districting in New Jersey and Hawaii.  Last, \textit{independent commissions} are appointed groups of unelected officials independent of the legislature, who have full authority for developing new redistricting plans and require bipartisan or non-partisan composition.  This is the sole legislative process for districting in Arizona, Idaho, California, and Washington at the congressional level \cite{Levitt}. \\

@@

  \section{Model}
  In this paper, we will conduct an analysis on the correlation between different measures of compactness and political bias, as a method for determining the overall asymmetry of a given political area.  Furthermore, we will examine the political processes involved in drawing those areas and compare the processes with the levels of gerrymandering exhibited.\\

@@

  \subsection {Computing Voter Efficiency}
  We calculate voter efficiency using data from a variety of sources.  The US Census Data, along with the Harvard Election Data Archive~\cite{heda} presents a great resource for Federal Election results.  State-by-state elections are best sourced from the respective state agencies that govern the election, such as the Secretary of State's office or Board of Elections.  To calculate voter efficiency, we first go back to the definition of wasted votes.  This measure of voter efficiency will only analyze two-party, single-winner elections.  Thus, for a given party $p$, their opposing party $o$ and a given district $i$, the number of wasted votes  if party $p$ wins ($V_{p} > V_{o}$)  is given by 
    \[ V_{wpi} = V_{pi} - \frac{V_{pi}+V_{oi}}{2}\]

    as all votes beyond 50\% are wasted. Otherwise, if party $p$ loses the district, then
    \[V_{wpi} = V_{pi}\]
    as all votes that do not go towards electing a candidate are wasted.\\

    While this calculates the wasted votes for an individual district, vote efficiency is calculated at the aggregate state level.  The vote efficiency $V_{Ep}$ for party $p$ is then calculated as follows:
     \[V_{Ep} = \frac{\sum_{i=1}^{n}(V_{wpi}-V_{woi})}{\sum_{i=1}^{n}(V_{pi}+V_{oi})}\]

     Stephanopolous and McGhee also propose that $V_{pi}+V_{oi}$ remains constant, as the size of any district in a given state cannot vary significantly, thus requiring equality.  Under this assumption, where the total votes in any given district are constant at $V_{t} = V_{pi}+V_{oi}$, and party $p$ wins $m$ of $n$ seats, the above equation simplifies to:
     \[V_{Ep} = (\frac{m}{n}-0.5) - 2*(\frac{\sum^{n}V_{pi}}{\sum^n{V_t}}-0.5)\]

  The first term, $(\frac{m}{n}-0.5)$ represents the ``seat share'' or the proportion of seats that were won by party $p$ in excess (or deficient) of half the seats.  The second term represents the ``vote share'' or the proportion of the vote that was won by party $p$ in excess of half. The 0.5 adjustment terms are there so that a positive $V_{Ep}$ indicates that the given party has an advantage due to partisan gerrymandering and a disadvantage with a negative $V_{Ep}$.  Furthermore, it represents the proportion of seats lost due to this disadvantage.\\

  Overall, the voter share/seat share metric is much easier to calculate, as nothing needs to be known about the individual districts, but rather only the totals of the overall election.  However, as we will expose in the analysis and discussion sections, the assumption that all districts have the same total votes does not hold well due to turnout variance and creates issues when analyzing the data.\\

@@

 \subsection{Redistricting Processes}
Taking in the states with at least 10 congressional districts, using both compactness and vote efficiency as a metric for measuring gerrymandering, we also find that legislative processes for redistricting produce more gerrymandered districts. The top half of states by each metric (Pennsylvania, Ohio, North Carolina, Michigan, and Virginia for both metrics, Texas for vote efficiency, and Illinois for compactness) all use legislative redistricting processes.  Likewise, three of the five states (California, Washington and New Jersey) in the bottom half of gerrymandering as measured by both metrics employ independent redistricting commissions or equally partisan political commissions.\\
Looking into this process further, we see this disparity even further in some of the most prominent of gerrymandered states -- Pennsylvania.  In Pennsylvania, the state legislature's Congressional plan produced an efficiency gap of nearly 24\%, or 4.27 seats.  However, the state house districts produced a seat gap of less than one -- out of over 200 seats.  This state house plan was drawn by an equally partisan political commission with representatives selected evenly by each party, thus forcing the state house districts to be drawn in a fair manner.  On the other hand, the Republican state legislature then subsequently forced through the plan that created the 4-seat bias, including a 26-24 vote in the state senate, with all Democrats and a few Republicans voting against the clearly gerrymandered plan.  The nature of such a biased legislative process clearly can lead to inequality in the system.  \\

@@

  \subsubsection{Voter Efficiency}
  As we have seen, voter efficiency works well as a measure for illegal gerrymandering, especially in concert with a good measure of compactness.  However, it does fall short in some situations.  For example, it does a poor job of measuring gerrymanders designed to protect incumbents.  In such a case, each party might waste a comparable number of votes, designed to protect the legislators still in office.  The voter efficiency gap would be low, yet districts would be predictable in outcome, protecting the incumbent.\\

  Similarly, races that are not contested can produce an extreme number of wasted votes for the winning party, with half of the total votes in that district being considered wasted.  While this may be indicative of partisan gerrymandering, it may also be reflective of a legislature people aren't as interested in.\\

@@

  We believe that this accurately captures the most egregious of gerrymandering, as for states with at least 3 congressional districts, a 20\% threshold limit or 2 seat gap creates a balance to test for gerrymandering in both small and large states.  For small states, this could mean a seat distribution changing from 3-0 to 2-1, which is small on the grand scale, but possibly offers better representation within that state. Likewise, it can catch large states, who can have a strong influence on the overall makeup of Congress.  Based on our data, it would invalidate districting plans in Florida, Michigan, New Jersey, North Carolina, Ohio, Pennsylvania, Texas, and Virginia on the basis that they have a gap larger than two seats, and Connecticut and Nebraska, on the basis of having greater than a 20\% efficiency gap. Providing this criteria will allow the courts to effectively decide on gerrymandering cases.\\

>>>>

  The very simplest metrics for compactness simply measure the length and width of the district as compared to that of a circle or rectangle of the same area.  There are several variations on this method, but they are all fairly crude.  Other metrics compare the area of the district to the area of the smallest circumscribing circle, or the largest inscribed circle, or to the convex hull of the district -- the smallest convex shape containing the entire district.  Niemi~et~al.~\cite{niemi} list a number of variations on these metrics.  An alternate method of computing dispersion is by computing the moment of inertia of the area of the district.  The concept is borrowed from physics, and is defined as
  \[I = \iint \left((x-x_0)^2 + (y-y_0)^2\right) dx\, dy \]
  where $(x_0, y_0)$ is the center of mass of the district (weighted by area).  More dispersed districts have higher moments of inertia, because the average square distance of a point in the district to the center is larger.  In order to ensure that the metric does not depend on the size of the state, and to make it range from 0 to 1 with 0 being the least compact, we in fact use the measure $A/\sqrt{2\pi I}$.  This metric seems to have been first proposed by Kaiser.~\cite{kaiser}

@@

  Lastly, since ultimately it is not the geography that matters, but the people that live in it, we can modify some of these metrics to take into account population.  Those that compare the area of the district to that of the smallest circumscribing circle or that of the convex hull of the district may simply compare population instead of area, as proposed by Hofeller and Grofman.~\cite{hofeller}.  Alternately, instead of computing the area-weighted moment of inertia, one can compute the population-weighted moment of inertia, which is defined as
  \[I_P = \iint \left((x-x_0)^2 + (y-y_0)^2\right) \rho\, dx\, dy \]
  where $\rho$ is the population density and $(x_0, y_0$ is the center of mass of the district, now weighted by population.  In practice, since we do not know the location of each individual in the district, we may simply assume that each person lives at the center of the census tabulation block in which they were counted.  Then the formula becomes
  \[I_P = \sum_{b} \left((x_b-x_0)^2 + (y_b-y_0)^2\right) P_b\]
  where the sum is taken over census tabulation blocks $b$, with $(x_b, y_b)$ the center of mass (weighted by area) of the census block, and $P_b$ its population.  In order to make the metric less dependent on population, we can use the metric $AP_d/(2\pi I_P)$; this does not necessarily normalize the metric to be between 0 and 1, but at least makes it independent of the total area and total population.  Using moment of population to measure district compactness was first proposed by Weaver.~\cite{weaver}

@@

    \subsection{Partisan Symmetry and Vote Efficiency}
 While geographic compactness is a widely-considered metric for compactness, others have focused on actual results for determining unacceptable redistricting plans.  Over the last few years, the courts have increasingly rejected a variety of rationales cited to claim illegal gerrymandering, including ``partisan intent'',  ``minority party entrenchment'', and ``lack of proportionality''.  However, in \textit{League of United Latin American Citizens (LULAC) v. Perry, Governor of Texas et al.}, (2006), the majority recognized that it would accept a lack of \textit{symmetry} as an acceptable argument for determining unconstitutional gerrymandering.

  A given plan has symmetry if ``similary-situated parties are treated equally'', thus expecting that an even split in votes across two parties would have an even split in seats. Furthermore, if a plaintiff challenging a redistricting plan can demonstrate partisan intent and asymmetry, it would be sufficient for the burden of proof to shift to the state.  Nevertheless, while acknowledging the viability of symmetry as a metric to determine a discriminatory impact on voters, the court failed to establish an accepted standard or test to determine illegal gerrymandering.  Existing measures at the time, as calculated by social scientists, used hypothetical models to predict how election outcomes would turn out, given certain shifts in the electorate and certain vote shares in a given election. The court rejected the notion of using hypothetical models for a standard of constitutionality. \cite{LULAC}

  With this in mind,  as an alternative to other existing metrics, two legal scholars recently proposed measuring \textit{vote efficiency} as a way to quantify partisan symmetry and partisan gerrymandering \cite{stephanopoulos}.  Vote efficiency is a measure of how well a given vote applied towards improving political representation.  In the case that a voter's desired candidate wins, any vote beyond what is necessary to win (in the simple case, 50\% + 1 vote) is wasted.  In the case when a voter's desired candidate loses, all votes for the losing candidate are wasted. Redistricting plans seeking a partisan bias will lead to fewer wasted votes for the favored party, while increasing wasted votes for the disadvantaged party.

@@

\subsection{Redistricting Processes}
Another key aspect when discussing gerrymandering is the political processes involved in redistricting.  The Constitution provides states with the ability to create their own rules in how districts are drawn, with the provisions that states will ensure balanced district populations, and that districts are redrawn at least every 10 years \cite{LULAC}.

Currently, there are five general types of processes, some of which overlap.  The first, and most common, is when redistricting is done by the \textit{legislature}.    The legislature proposes plans and seeks to pass a bill for approval.  Sometimes, this may require a supermajority, or may be veto-proof, but typically requires simple majority approval from both the upper and lower houses and from the governor.  The second is the \textit{advisory commission} model, where a group of people, usually appointed officials, gather together to make non-binding recommendations to the legislature.  The third is the \textit{backup commission}, where failure of the legislature to pass a new plan triggers the activation of a predetermined committee to create an alternative.  The fourth is a \textit{political commission}, where elected officials often serve as members of a redistricting commission.  This is the sole legislative process for districting in New Jersey and Hawaii.  Last, \textit{independent commissions} are appointed groups of unelected officials independent of the legislature, who have full authority for developing new redistricting plans and require bipartisan or non-partisan composition.  This is the sole legislative process for districting in Arizona, Idaho, California, and Washington at the congressional level \cite{Levitt}.

@@

  \section{Model}
  In this paper, we will conduct an analysis on the correlation between different measures of compactness and political bias, as a method for determining the overall asymmetry of a given political area.  Furthermore, we will examine the political processes involved in drawing those areas and compare the processes with the levels of gerrymandering exhibited.

@@

  \subsection {Computing Voter Efficiency}
  We calculate voter efficiency using data from a variety of sources.  The US Census Data, along with the Harvard Election Data Archive~\cite{heda} presents a great resource for Federal Election results.  State-by-state elections are best sourced from the respective state agencies that govern the election, such as the Secretary of State's office or Board of Elections.  To calculate voter efficiency, we first go back to the definition of wasted votes.  This measure of voter efficiency will only analyze two-party, single-winner elections.  Thus, for a given party $p$, their opposing party $o$ and a given district $i$, the number of wasted votes  if party $p$ wins ($V_{p} > V_{o}$)  is given by
    \[ V_{wpi} = V_{pi} - \frac{V_{pi}+V_{oi}}{2}\]
  as all votes beyond 50\% are wasted. Otherwise, if party $p$ loses the district, then
    \[V_{wpi} = V_{pi}\]
  as all votes that do not go towards electing a candidate are wasted.

    While this calculates the wasted votes for an individual district, vote efficiency is calculated at the aggregate state level.  The vote efficiency $V_{Ep}$ for party $p$ is then calculated as follows:
     \[V_{Ep} = \frac{\sum_{i=1}^{n}\left(V_{wpi}-V_{woi}\right)}{\sum_{i=1}^{n}\left(V_{pi}+V_{oi}\right)}\]

     Stephanopolous and McGhee also propose that $V_{pi}+V_{oi}$ remains constant, as the size of any district in a given state cannot vary significantly, thus requiring equality.  Under this assumption, where the total votes in any given district are constant at $V_{t} = V_{pi}+V_{oi}$, and party $p$ wins $m$ of $n$ seats, the above equation simplifies to:
     \[V_{Ep} = \left(\frac{m}{n}-0.5\right) - 2*\left(\frac{\sum^{n}V_{pi}}{\sum^n{V_t}}-0.5\right)\]

  The first term, $\left(\frac{m}{n}-0.5\right)$ represents the ``seat share'' or the proportion of seats that were won by party $p$ in excess (or deficient) of half the seats.  The second term represents the ``vote share'' or the proportion of the vote that was won by party $p$ in excess of half. The 0.5 adjustment terms are there so that a positive $V_{Ep}$ indicates that the given party has an advantage due to partisan gerrymandering and a disadvantage with a negative $V_{Ep}$.  Furthermore, it represents the proportion of seats lost due to this disadvantage.

  Overall, the voter share/seat share metric is much easier to calculate, as nothing needs to be known about the individual districts, but rather only the totals of the overall election.  However, as we will expose in the analysis and discussion sections, the assumption that all districts have the same total votes does not hold well due to turnout variance and creates issues when analyzing the data.

@@

 \subsection{Redistricting Processes}
Taking in the states with at least 10 congressional districts, using both compactness and vote efficiency as a metric for measuring gerrymandering, we also find that legislative processes for redistricting produce more gerrymandered districts. The top half of states by each metric (Pennsylvania, Ohio, North Carolina, Michigan, and Virginia for both metrics, Texas for vote efficiency, and Illinois for compactness) all use legislative redistricting processes.  Likewise, three of the five states (California, Washington and New Jersey) in the bottom half of gerrymandering as measured by both metrics employ independent redistricting commissions or equally partisan political commissions.
Looking into this process further, we see this disparity even further in some of the most prominent of gerrymandered states -- Pennsylvania.  In Pennsylvania, the state legislature's Congressional plan produced an efficiency gap of nearly 24\%, or 4.27 seats.  However, the state house districts produced a seat gap of less than one -- out of over 200 seats.  This state house plan was drawn by an equally partisan political commission with representatives selected evenly by each party, thus forcing the state house districts to be drawn in a fair manner.  On the other hand, the Republican state legislature then subsequently forced through the plan that created the 4-seat bias, including a 26-24 vote in the state senate, with all Democrats and a few Republicans voting against the clearly gerrymandered plan.  The nature of such a biased legislative process clearly can lead to inequality in the system.

@@

  \subsubsection{Voter Efficiency}
  As we have seen, voter efficiency works well as a measure for illegal gerrymandering, especially in concert with a good measure of compactness.  However, it does fall short in some situations.  For example, it does a poor job of measuring gerrymanders designed to protect incumbents.  In such a case, each party might waste a comparable number of votes, designed to protect the legislators still in office.  The voter efficiency gap would be low, yet districts would be predictable in outcome, protecting the incumbent.

  Similarly, races that are not contested can produce an extreme number of wasted votes for the winning party, with half of the total votes in that district being considered wasted.  While this may be indicative of partisan gerrymandering, it may also be reflective of a legislature people aren't as interested in.

@@

  We believe that this accurately captures the most egregious of gerrymandering, as for states with at least 3 congressional districts, a 20\% threshold limit or 2 seat gap creates a balance to test for gerrymandering in both small and large states.  For small states, this could mean a seat distribution changing from 3-0 to 2-1, which is small on the grand scale, but possibly offers better representation within that state. Likewise, it can catch large states, who can have a strong influence on the overall makeup of Congress.  Based on our data, it would invalidate districting plans in Florida, Michigan, New Jersey, North Carolina, Ohio, Pennsylvania, Texas, and Virginia on the basis that they have a gap larger than two seats, and Connecticut and Nebraska, on the basis of having greater than a 20\% efficiency gap. Providing this criteria will allow the courts to effectively decide on gerrymandering cases.
